---
title: "UNICEF Coverage Report: Antenatal Care and Skilled Birth Attendance (2018–2022)"
author: "UNICEF D\\&A Education Team"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cerulean  
    highlight: tango
    df_print: paged
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
  word_document: default
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
library(plotly)
library(scales)
library(knitr)
library(kableExtra)
```

```{r load-data}

df_group <- readRDS(file.path(dir_data_output, "coverage_results_by_group.rds"))
df_country <- readRDS(file.path(dir_data_output, "coverage_results_by_country.rds"))
df_summary <- readRDS(file.path(dir_data_output, "coverage_summary_group.rds"))
df_top10_combined <- readRDS(file.path(dir_data_output, "top10_combined_coverage.rds"))
df_top10 <- readRDS(file.path(dir_data_output, "top10_anc4.rds"))
df_long <- readRDS(file.path(dir_data_output, "coverage_long_data.rds"))

# Update indicator labels
df_long <- df_long %>%
  mutate(indicator = recode(indicator,
    anc4_coverage = "ANC4: 4+ Antenatal Care Visits",
    sba_coverage  = "SBA: Skilled Birth Attendance"
  ))
```

> **About this Submission**\
> This technical report supports the application to the following UNICEF D&A consultancy positions:\
> – Learning and Skills Data Analyst Consultant (#581598)\
> – Household Survey Data Analyst Consultant (#581656)\
> – Administrative Data Analyst Consultant (#581696)\
> – Microdata Harmonization Consultant (#581699)

# 1. Context, Objectives and Methodology

------------------------------------------------------------------------

> **About this Report**\
> This technical note summarizes coverage of maternal health services across UNICEF-focus countries. Countries are classified as **on-track** or **off-track** based on under-five mortality status (U5MR), aligned with **SDG 3.2**. The indicators analyzed — ANC4 and SBA — are key components of the maternal care continuum and contribute to the achievement of **SDG 3.1**.

------------------------------------------------------------------------

### Context

Reducing **under-five mortality** remains a global priority — particularly in countries where maternal health services are limited or inequitable. To assess readiness and identify service gaps, UNICEF monitors two critical indicators of maternal care:

-   **ANC4**: Percentage of women (aged 15–49) receiving at least **four antenatal care visits**.
-   **SBA**: Percentage of births attended by **skilled health personnel**.

These indicators reflect **early engagement** in pregnancy and **safe delivery**, both essential to reducing maternal and neonatal mortality. Together, they form a core part of the continuum of care.

To contextualize coverage performance, countries are grouped using the latest **under-five mortality rate (U5MR) classification**:

-   **On-track**: U5MR is *achieved* or *on-track* to meet **SDG target 3.2**.
-   **Off-track**: U5MR status is *acceleration needed*.

This classification provides a framework to assess how service coverage aligns with child survival progress and helps inform areas where additional attention may be needed to meet **SDG target 3.2** on ending preventable child deaths.

------------------------------------------------------------------------

### Methodology

#### 1. Data Preparation

-   **ANC4** and **SBA** coverage data were retrieved from the\
    [UNICEF Global Data Repository](https://data.unicef.org/resources/data_explorer/unicef_f/?ag=UNICEF&df=GLOBAL_DATAFLOW&ver=1.0&dq=.MNCH_ANC4+MNCH_SAB.&startPeriod=2018&endPeriod=2022)\
    for all available countries between **2018 and 2022**.
-   For each country, the **most recent coverage estimate** within this period was retained.
-   **Under-five mortality classification** (on-track / off-track) was assigned using the\
    file `On-track and off-track countries.xlsx`, based on U5MR status:
    -   *On-track* = “achieved” or “on-track”
    -   *Off-track* = “acceleration needed”
-   All datasets were merged using standardized **ISO3 country codes**.

#### 2. Population-Weighted Averages

-   **Birth projections for 2022** were sourced from the UN World Population Prospects:\
    `WPP2022_GEN_F01_DEMOGRAPHIC_INDICATORS_COMPACT_REV1.xlsx`.
-   For each group (on-track and off-track), **weighted means** were computed for ANC4 and SBA using birth counts as weights.
-   Countries with missing coverage data were excluded from group-level weighted statistics.

------------------------------------------------------------------------

> **Data Sources and Limitations**\
> - **Date of extraction and processing**: July 2025\
> - **Inputs**:\
> - ANC4 and SBA indicators from the [UNICEF Global Data Repository](https://data.unicef.org/resources/data_explorer/unicef_f/?ag=UNICEF&df=GLOBAL_DATAFLOW&ver=1.0&dq=.MNCH_ANC4+MNCH_SAB.&startPeriod=2018&endPeriod=2022)\
> - Birth projections from the *UN World Population Prospects 2022*\
> - U5MR groupings from UNICEF’s official classification
>
> Several countries were **excluded or only partially included** due to missing data (e.g., births, U5MR status, ANC4 or SBA coverage).\
> These limitations were accounted for in the weighted analysis and are examined in detail in **Section 2.2 on data completeness**.

------------------------------------------------------------------------

# 2. Analysis of Coverage and Gaps

This section presents group-level and country-level results for ANC4 and SBA coverage, highlighting disparities, data quality issues, and patterns relevant for maternal health programming.

------------------------------------------------------------------------

### 2.1 Summary coverage statistics by group

```{r}
df_summary %>%
  mutate(
    `Avg. ANC4` = round(anc4_mean, 1),
    `Avg. SBA` = round(sba_mean, 1),
    `Missing ANC4` = anc4_missing,
    `Missing SBA` = sba_missing
  ) %>%
  select(Group = group, `# Countries` = n_countries, `Avg. ANC4`, `Avg. SBA`, `Missing ANC4`, `Missing SBA`) %>%
  kable(caption = "Coverage Summary by Group") %>%
  kable_styling(full_width = F, bootstrap_options = c('striped', 'hover'))
```

**Interpretation**\
This table shows the average ANC4 and SBA coverage for on-track and off-track countries, along with the number of countries and missing values.

Key insights include: - SBA coverage is consistently high across both groups. - ANC4 coverage is lower and more variable, especially in off-track countries. - Over 60% of on-track countries have no ANC4 data, limiting comparability.

------------------------------------------------------------------------

### 2.2 Data completeness by indicator

```{r data-completeness-stackedbar}

# % of countries excluded from final analysis
pct_excluded <- round(
  100 * (excluded_countries %>% distinct(iso3) %>% nrow()) /
    ((excluded_countries %>% distinct(iso3) %>% nrow()) +
       (df_merge %>% distinct(iso3) %>% nrow())),
  1
)


df_completeness <- df_country %>%
  pivot_longer(cols = c(anc4_weighted, sba_weighted), names_to = "indicator", values_to = "value") %>%
  mutate(
    indicator = recode(indicator,
                       anc4_weighted = "ANC4",
                       sba_weighted  = "SBA"),
    completeness = ifelse(is.na(value), "Missing", "Available")
  ) %>%
  count(group, indicator, completeness) %>%
  group_by(group, indicator) %>%
  mutate(pct = 100 * n / sum(n))

ggplot(df_completeness, aes(x = indicator, y = pct, fill = completeness)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ group) +
  scale_fill_manual(
    values = c("Missing" = "#DBB0AD", "Available" = "#ADCEDB")
  ) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(
    title = "Data Completeness for ANC4 and SBA by Group",
    y = "Percentage of Countries",
    x = "Indicator",
    fill = "Data Availability"
  ) +
  theme_minimal(base_size = 13)
```

**Interpretation**\
- Over 60% of on-track countries lack ANC4 data, weakening group-level reliability.\
- SBA data is nearly complete, enabling stronger comparisons.\
- *`r pct_excluded`%* of countries were fully excluded due to missing group, births, or both indicators.\
- Others were partially included, lacking ANC4 but contributing SBA data — highlighting antenatal reporting gaps.\
- The table below outlines missing data patterns to guide data system improvements.

```{r exclusion-summary-table}
excluded_summary %>%
  kable(caption = sprintf("Missing data patterns")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

------------------------------------------------------------------------

### 2.3 Country-Level coverage: interactive view

```{r}
plot_interactive <- ggplot(df_long, aes(x = reorder(country, coverage), y = coverage,
                         color = group, text = paste0(
                           "Country: ", country, "<br>",
                           "Group: ", group, "<br>",
                           "Indicator: ", indicator, "<br>",
                           "Coverage: ", round(coverage, 1), "%"
                         ))) +
  geom_point(size = 2.3, alpha = 0.8) +
  facet_wrap(~ indicator, scales = "free_x") +
  scale_y_continuous(limits = c(0, 100), labels = label_percent(scale = 1)) +
  labs(
    title = "Coverage by Country, Group and Indicator",
    x = "Country",
    y = "Coverage (%)",
    color = "Group"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_blank())

ggplotly(plot_interactive, tooltip = "text")
```

**Interpretation**\
This graph highlights the contrast between **ANC4** and **SBA** coverage by country group:

-   **SBA coverage** is consistently high, with most countries reaching or exceeding **80%**.
-   **ANC4 coverage**, however, is more variable — particularly in *off-track* countries, where it often falls **below 60%**.

➡️ *See Section 2.4 for summary distributions.*

> **Note**: Best viewed in HTML for tooltip interactivity.

------------------------------------------------------------------------

### 2.4 Coverage distribution by group

```{r}
ggplot(df_long, aes(x = indicator, y = coverage, fill = group)) +
  geom_boxplot(width = 0.4, outlier.shape = NA, position = position_dodge(0.8), alpha = 0.5) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 1) +
  scale_y_continuous(labels = label_percent(scale = 1), limits = c(0, 100)) +
  labs(
    title = "Coverage Distribution by Indicator and Group",
    x = "Indicator",
    y = "Coverage (%)",
    fill = "Group"
  ) +
  theme_minimal(base_size = 13)
```

**Interpretation**\
This distribution plot shows the **range and central tendency** of ANC4 and SBA coverage by country group. The **median SBA is high** for both groups, while **ANC4 coverage is lower and more dispersed**, especially in off-track settings. On-track countries appear to perform slightly better in ANC4, but many are missing data, as noted earlier.

The wider spread in ANC4 for off-track countries also highlights **greater inequality in antenatal care access**, pointing to **potential structural or systemic challenges** in reaching women early in pregnancy.

------------------------------------------------------------------------

### 2.5 ANC4 vs SBA Scatterplot

```{r scatter-anc4-sba}
ggplot(df_country, aes(
  x = anc4_weighted,
  y = sba_weighted,
  color = group,
  label = country
)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "grey50") +
  scale_x_continuous(name = "ANC4 Coverage (%)", limits = c(0, 100), labels = label_percent(scale = 1)) +
  scale_y_continuous(name = "SBA Coverage (%)", limits = c(0, 100), labels = label_percent(scale = 1)) +
  labs(
    title = "ANC4 vs SBA Coverage by Country",
    subtitle = "Each point represents a country",
    color = "Group"
  ) +
  theme_minimal(base_size = 13)
```

**Interpretation**\
This scatterplot highlights the relationship between ANC4 and SBA coverage across countries. While many countries show a positive correlation, a significant cluster of off-track countries exhibit high SBA coverage but low ANC4, indicating that women are receiving delivery care without adequate antenatal follow-up.

------------------------------------------------------------------------

> **Key Summary Insight (applies across Sections 2.1–2.5):**\
> **SBA coverage is consistently strong**, indicating widespread access to delivery care. In contrast, **ANC4 coverage remains limited or missing**, especially in off-track countries and many on-track countries. This imbalance reveals critical gaps in early maternal engagement and system continuity.

------------------------------------------------------------------------

### 2.6 Countries with lowest ANC4 coverage

```{r}
df_top10_combined %>%
  rename(
    `ANC4 Coverage (%)` = anc4_coverage,
    `SBA Coverage (%)`  = sba_coverage,
    Group = group,
    Country = country
  ) %>%
  kable(caption = "Top 10 Countries with Lowest Combined ANC4 and SBA Coverage") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c('striped', 'hover'))
```

> These countries face compounded vulnerabilities in both antenatal and delivery services. Addressing gaps in the full maternal health continuum is essential for reducing preventable maternal and neonatal deaths.

------------------------------------------------------------------------

# 3. Interpretation

The results highlights **strong skilled birth attendance (SBA)** across all countries, but also reveals **significant challenges with antenatal care coverage (ANC4)**, particularly in **off-track countries** and in **on-track countries** where data is missing or incomplete. This imbalance raises concerns about the **continuity of maternal care**, a key element for achieving **SDG targets 3.1 and 3.2**.

#### **Coverage comparison by country group**

| **Metric** | **On-Track Countries** | **Off-Track Countries** |
|--------------------|------------------------|----------------------------|
| **SBA coverage** | High, with little variation | Also high and relatively stable |
| **ANC4 coverage** | More heterogeneous | Often low, typically \< 60% |
| **ANC4 data completeness** | Poor: \~60% of countries missing | Good: Nearly all countries reported |

> These results suggest that while many women receive skilled support during delivery, antenatal engagement remains limited in several contexts — either due to insufficient coverage or the absence of reliable data.

### **Why is ANC4 Low or Missing?**

ANC4 is inherently **more difficult to capture and deliver** than SBA. Unlike delivery, which is a single event often occurring in a facility, ANC4 requires **multiple visits**, spread over months. Several factors contribute to low uptake or missing data:

-   **Economic barriers** (e.g., transport costs, informal fees)
-   **Time constraints** (especially for women with household or work responsibilities)
-   **Distance and infrastructure**
-   **Lack of awareness or trust in health systems**

> In contrast, SBA, a one-time event, is easier to track and often better integrated into health systems.

Several of the **lowest-performing countries** , including *Somalia, South Sudan, Niger, Central African Republic*, and *Afghanistan* — report **ANC4 and SBA coverage both below 50%**. These countries represent **critical gaps in maternal care** that affect both early and delivery-stage interventions. Others, such as *Senegal* and *Mauritania*, show **moderate SBA coverage but extremely low ANC4 uptake**, pointing to a **breakdown in continuity of care** during pregnancy. All countries in this list are **off-track** in under-five mortality, reinforcing the urgency of targeted action.

------------------------------------------------------------------------

# 4 Strategic Recommendations

#### **For On-Track Countries**

-   **Close the ANC4 data gap**:
    -   Improve antenatal data availability through routine health information systems and survey coverage.
    -   When unavailable, triangulate ANC4 estimates using facility, insurance, or digital health records.
-   **Ensure quality beyond SBA**:
    -   Validate success with maternal care: High SBA alone isn’t sufficient. Without ANC4, risks remain high. Ensure “on-track” status includes quality and continuity of care.

#### **For Off-Track Countries**

-   **Invest in antenatal access**:
    -   Address **geographic, economic, and sociocultural barriers** to early pregnancy care.\
    -   Use **mobile clinics, community health workers**, and local outreach to reach underserved women.
    -   *Note: Countries with both low ANC4 and SBA (e.g., **Somalia**, **South Sudan**, **Niger**, **CAR**) require foundational investment in access and coverage.*
-   **Correct the care imbalance**:
    -   In countries with high SBA but low ANC4 (e.g., **Senegal**, **Mauritania**), encourage early pregnancy contact through targeted interventions and communication strategies.

#### **For All Countries**

-   **Enhance equity and targeting**:
    -   Disaggregate ANC4/SBA coverage by **age, location, wealth, and parity** to uncover hidden gaps.
-   **Institutionalize annual monitoring**:
    -   Align ANC4/SBA tracking with DHS/MICS cycles and integrate indicators into national dashboards.
-   **Commission deeper research**
    -   A **targeted study is needed** to understand low ANC4 uptake, combining **demand-side factors** (awareness, distance, cost) with **supply-side barriers** (health system reach, data gaps).

------------------------------------------------------------------------

# 5. Sources

-   WHO Indicator Metadata:
    -   [ANC4](https://www.who.int/data/gho/indicator-metadata-registry/imr-details/80)
    -   [SBA](https://data.who.int/indicators/i/F835E3B/1772666)
-   [UNICEF Under-Five Mortality (U5MR) Overview](https://data.unicef.org/topic/child-survival/under-five-mortality/) – used to define on-track and off-track country groupings.
-   UN SDG Indicator Metadata:
    -   [SDG 3.1.1 – Maternal Mortality](https://unstats.un.org/sdgs/metadata/files/Metadata-03-01-01.pdf)
    -   [SDG 3.1.2 – Skilled Birth Attendance (SBA)](https://unstats.un.org/sdgs/metadata/files/Metadata-03-01-02.pdf)
    -   [SDG 3.2.1 – Under-Five Mortality](https://unstats.un.org/sdgs/metadata/files/Metadata-03-02-01.pdf)
    -   [SDG 3.2.2 – Neonatal Mortality](https://unstats.un.org/sdgs/metadata/files/Metadata-03-02-02.pdf)
-   Data derived from analytical script `02_analysis.R` (consultancy assessment deliverable)
